Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-09-16 writeups/research_proposal/propsal.qmd

---
title: "Comparison of MCMC algorithms in Stochastic Volatility Models"
author: 
    - "By Benjamin Wee" 
    - "Supervised by Catherine Forbes and Lauren Kennedy"
# abstract: "The objective of this research is to validate and compare the computational methods used to estimate Bayesian stochastic volatility models. Specifically, Simulation Based Calibration will be used to identify biases and errors arising from the results of Markov Chain Monte Carlo (MCMC) algorithms. The source of any miscalibration include, but are not limited to, estimation strategies, choice of sampler, model specifications and coding mistakes. This study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo for estimating these class of models. Results and diagnostics from these simulations will help identify strengths and limitations that arise from these approaches."
format:
  pdf:
    fig-pos: H
    include-in-header:
        text: |
          \usepackage{lipsum}
          \usepackage{setspace}
          \onehalfspacing
          \linespread{1.5}
fontsize: 12pt
geometry: margin=1in
mathspec: true
---

@@

## 2.1 Stochastic Volatility
Stochastic volatility models are used in financial econometrics to model the variance or volatility of returns in a financial instrument. The return of a financial asset often exhibits pattterns in the variance (heteroskedasticity) which can be modelled to improve prediction and estimates of risk or tail events. Unlike other volatility models such as ARCH or GARCH, these stochastic volatility models treat the variance as a random variable. Furthermore, these models have a state space representation since the variance is treated as a latent stochastic process. State space models can be difficult to estimate since there is a latent state for every data observation in the time series, which means there are more parameters than data points. These models can be estimated using both frequent tools but this research will focus on Bayesian estimation strategies.

Kim, Sherpherd and Chib (1998) estimate a discrete time, univariate stochastic volatility model using Bayesian methods which is outlined below. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is its variance. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks.

@@

### Estimation strategy
Kim, Shephard and Chib's sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother[^1] (de Jong and Shephard, 1995).

@@

@@

## 2.2 Hamiltoninan Monte Carlo
Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature (Duane et al. 1987) before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks (Nael, 1995) and statistical computing (Neal 2011). The algorithm has since become widely available through open source development projects such as Stan (Stan Development Team, 2023) and PyMC (2023).

The key innovation of Hamiltonian Monte Carlo is using the gradients of the target posterior distribution to generate an efficient path for the sampler to explore. Unlike Random Walk Metropolis Hastings, it takes advantage of the geometry of the posterior to determine its next step. A comprehensive explanation of the sampler is beyond the scope of this research and can be found in the above references.

The Stan programming language's implementation of Hamiltonian Monte Carlo will be used for this study. Stan's default algorithm, the No-U-Turn Sampler (Hoffman & Gelman 2014), allows for direct sampling of the specified stochastic volatility model. Hamiltonian Monte Carlo allows for sampling of the generative model and can flexibly handle complicated likelihood functions. This approach will use the same priors as specified in the Gaussian mixture approximation. However, Hamiltonian Monte Carlo does not sample directly from the analytical form of conjugate or conditionally conjugate distribution, so different priors can be chosen if necessary.

## 2.3 Research contribution
The goal of this research is to determine which algorithm is better at estimating stochastic volatility models. There are clear differences between Kim Sherphard and Chib's estimation strategy and Hamiltoninan Monte Carlo. My main contribution is to compare these algorithms and determine which of these sampling strategies is better based on whether the algorithm returns calibrated posterior estimates and the efficiency of the MCMC sampler.

@@

## 3.1 Design
Simulation Based Calibration (SBC) is a simulation technique that will help check the calibration of posterior estimates generated by different MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples estimated from the simulated data. Let $\theta$ be a parameter and $y$ represent the dataset. Start with draw a sample from the prior distribution:
$$
\begin{aligned}
\theta^{sim} \sim \pi(\theta)
\end{aligned}
$$

Generate a dataset given by the prior draws.

@@

Then take draws from the posterior distribution generated by a MCMC algorithm or estimation strategy (Hamiltonian Monte Carlo or Gaussian approximation in our case).

@@

A key result from this procedure is that the posterior sample $\{\theta_1,\dots , \theta_{L}\}$ will share the same distribution as the prior samples $\theta^{sim}$. This is implied by the following expression:

@@

That is, the posterior averaged over the joint distribution follows the same distribution as the prior. The procedure of generating posterior samples implicitly performs this integral since the expression on the right of the integral is proportional to the prior density. Therefore, any deviation of the posterior samples from the prior distribution suggests an error - either in computation, code or overall analysis.

@@

Since the distribution of the posterior samples follows the prior distribution, the rank statistic for a given parameter and simulation follows a discrete uniform distribution[^2].

[^2]: Proof of this result in Talts et al 2018.

@@

This completes one iteration of SBC. To complete the algorithm, multiple iterations are run and the rank statistics are calculated for each parameter. The resulting rank statistics are compared to the discrete uniform distribution to determine if any problematic features or errors exist.

If the computation is well calibrated and the samples follow a discrete uniform distribution, then the posterior (credible) intervals will have sufficient coverage. That is, for any percentage interval selected (for example 90%) then there is 90\% chance that any $\theta^{sim}$ falls within this interval. In other words, a Bayesian analysis and computation is well calibrated if the 90\% interval contains the true parameter in 90\% interations of the SBC procedure.

## 3.2 Implementation
SBC for Hamiltonian Monte Carlo and the Gaussian approximation are performed with simulated dataset sizes of 1000 simulation iterations, 1000 observations, and post warmup/burnin samples of 999 (which gives 1000 possible rank statistics). This algorithm is summarised below:

@@

## 3.3 Metrics
The key metrics and diagnostics to compare the performance of these samplers is the effective sample size, histograms of rank statistics and chi-squared test statistics.

Effecitve sample size (ESS) measures the efficiency of the MCMC sampler. It calculates the number of (effectively) independent draws from the posterior draws generated by a Markov chain. A poor ESS can arise from high autocorrelation in the Markov chain which leads to highly dependent samples. If it requires long MCMC chains to get to a desired ESS, then this is considered inefficient since it takes more resources to get a representative sample of the target distribution.

Histograms of rank statistics are a visual diagnostic to determine if the distribution of ranks follows a discrete uniform distribution. As described, a calibrated algorithm implies rank statistics which are randomly drawn from a uniform. A weakness of this approach is there are at least as many parameters as there data points. Since there are 1000 observations per dataset, this means there are 1003 parameters to estimate.

An alternative to checking for uniformity of all the parameters is to calculate the chi squared statistics for the counts in each bin. Let $b_j$ be the number of counts and $e_j$ the expected count in bin $j$. Then the chi squared statistic is given by:

@@

The chi squared statistics provide a more succinct summary of the overall performance of the algorithm.

@@

# 4. Preliminary Results
Figure 1 shows the distribution of rank statistics for Hamiltonian Monte Carlo, with the horizontal black line at 50 representing the discrete uniform distribution. The preliminary figures focus on the static parameters and arbitrarily chosen states (since there is one state parameter per data observation). Phi and sigma_sqd look relatively uniform. However mu and h.995 (the 995th state parameter) have some lumpiness which may suggest a lack of calibration.

@@

One of the drawbacks of SBC is that any deviation from uniform from a calibrated analysis maybe due to finite samples. Figure 2 shows more consistent uniform behaviour from all parameters after increasing the number of iterations from 1000 to 5000. This suggests that Hamiltonian Monte Carlo and overall analysis is well calibrated.

@@

Figure 3 shows results for the Gaussian approximation. The distribution of the rank statistics deviates heavily from the expected discrete uniform distribution, with various peaks at each end of the histogram as well as peaks in the middle. This lack of calibration is expected. The Gaussian approximation only approximates the log chi squared distribution, whereas Hamiltonian Monte Carlo samples directly from the same likelihood as the data generating process.

This result may also be due to the small number of posterior samples drawn from the model. The MCMC sampler may not have converged yet which may explain the deviations from uniformity. A follow up step is to compare the effective sample sizes (ESS) of the two methods and to increase the length of the MCMC chain to ensure the ESS is approximately the same before comparing histograms.

![Distribution of parameter ranks for KSC sampling strategy (1000 iterations)](../../simulation_output/sbc_cp_ksc_model_cp_dgf_r1/static_state_hist.png){width=75%}

@@

Kim, Sherpherd and Chib apply a Metropolis Hastings reweighting to the posterior samples to correct for the approximation of the error distribution. According to the authors, the correction ensures that samples are from the exact posterior density. The next step is to perform this correction and evaluate the SBC results. I'd expect the Gaussian mixture approximation diagnostics to improve after this correction.

@@

The parameterisation of a model can affect the performance of a sampler. This can cause significant problems in getting unbiased MCMC samples for complex models. An example of this is Neal's funnel (Neal 2003) which arises in hierarchical models which causes performance issues with the Hamiltonian Monte Carlo sampler. Rewriting the model with non centered parameterisation significantly improves the performance of MCMC. Another consideration for this research is to consider both centered and non centered parametersations of the stochastic volatility model to see if there are any improvements to the simulation diagnostics.

@@

![Model matrix](model_table.png)

@@

Week 8: Finish off any simulation or modeling not yet complete (listed in above matrix)

Week 9: Write up any other results and have a first draft ready for feedback

@@

>>>>

---
title: "Comparison of MCMC algorithms in Stochastic Volatility Models"
author: 
    - "By Benjamin Wee (32921020)" 
    - "Supervised by Catherine Forbes and Lauren Kennedy"
# abstract: "The objective of this research is to validate and compare the computational methods used to estimate Bayesian stochastic volatility models. Specifically, Simulation Based Calibration will be used to identify biases and errors arising from the results of Markov Chain Monte Carlo (MCMC) algorithms. The source of any miscalibration include, but are not limited to, estimation strategies, choice of sampler, model specifications and coding mistakes. This study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo for estimating these class of models. Results and diagnostics from these simulations will help identify strengths and limitations that arise from these approaches."
format:
  pdf:
    fig-pos: H
    include-in-header:
        text: |
          \usepackage{lipsum}
          \usepackage{setspace}
          \onehalfspacing
          \linespread{1.5}
fontsize: 12pt
geometry: margin=1in
mathspec: true
---

@@

## 2.1 Stochastic Volatility
Stochastic volatility models are used in financial econometrics to model the variance or volatility of returns in a financial instrument. The return of a financial asset often exhibits patterns in the variance which can be modeled to improve prediction and estimates of risk or tail events. Unlike other volatility models such as ARCH or GARCH, stochastic volatility models treat the variance as a random variable. Furthermore, these models have a state space representation since the variance is expressed as a latent stochastic process. State space models can be difficult to estimate since there is a latent state for every data observation in the time series, which means there are often more parameters than data points. These models can be estimated using frequent frameworks but this research will focus on Bayesian estimation strategies.

Kim, Shephard and Chib (1998) estimate a discrete time, univariate stochastic volatility model using Bayesian methods which is outlined below. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

### Estimation strategy
Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother[^1] (de Jong and Shephard, 1995).

@@

Where $z_t = log(\epsilon_t^2)$ follows a log chi-squared distribution with mean -1.2704 and variance 4.93. The relationship between $y_t$ and $h_t$ is now linear; however, the error is not Gaussian. Since it is not simple to sample from this parameterisation of the model, KSC use a mixture of Gaussians to **approximate** the first 4 moments of the log chi squared distribution. This is defined by:

@@

The model can be sampled via the Kalman Filter and simulation smoother since the model is now linear and conditionally gaussian. The static parameters $\mu$ and $\sigma^2$ are sampled directly from their conjugate posterior distributions whereas $\phi$ is sampled via a Metropolis Hastings accept/reject procedure. The details can be around in Appendix B.

## 2.2 Hamiltoninan Monte Carlo
Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature (Duane et al. 1987) before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks (Neal, 1995) and statistical computing (Neal 2011). The algorithm has since become widely available through open source development projects such as Stan (Stan Development Team, 2023) and PyMC (2023).

The key innovation of Hamiltonian Monte Carlo is using the gradients of the target posterior distribution to generate an efficient path for the sampler to explore. Unlike Random Walk Metropolis Hastings, it takes advantage of the geometry of the posterior to determine its next proposal step. A comprehensive explanation of the sampler is beyond the scope of this research and can be found in the above references.

The Stan programming language's implementation of Hamiltonian Monte Carlo will be used for this study. Stan's default algorithm, the No-U-Turn Sampler (Hoffman & Gelman 2014), allows for direct sampling of the specified stochastic volatility model. Hamiltonian Monte Carlo allows for sampling of the generative model and can flexibly handle complicated likelihood functions. This approach will also use the same priors as specified in the Gaussian mixture approximation.

## 2.3 Research contribution
The goal of this research is to determine which algorithm is better at estimating the discrete time stochastic volatility model (conditional on the same priors). My main contribution is to design a simulation study that compares these algorithms and determines which sampling strategy is better based on the calibration of posterior estimates and the efficiency of the MCMC sampler.

@@

## 3.1 Design
Simulation Based Calibration (SBC) is a technique that checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples conditional on the simulated data.

To illustrate this procedure, let $\theta$ be a parameter and $y$ represent the dataset. Start with a single draw from the prior distribution:
$$
\begin{aligned}
\theta^{sim} \sim \pi(\theta)
\end{aligned}
$$

Generate a dataset given by the prior draw.

@@

Then take draws from the posterior distribution generated by a MCMC algorithm or estimation strategy (Hamiltonian Monte Carlo or KSC) conditional on this dataset.

@@

A key result is that the posterior sample $\{\theta_1,\dots , \theta_{L}\}$ will share the same distribution as the prior samples $\theta^{sim}$. This is implied by the following expression:

@@

That is, the posterior averaged over the joint distribution follows the same distribution as the prior. The procedure of generating posterior samples implicitly performs this integral since the expression on the right of the integral is proportional to the prior density. Therefore, any deviation of the posterior samples from the prior distribution suggests that the sampling methodology is not producing the correct posteriors.

<!-- n error - either in computation, code or overall analysis.  -->

@@

If the posterior samples follows the prior distribution, the rank statistic for a given parameter follows a discrete uniform distribution[^2]. The rank statistic is defined as:

[^2]: Proof of this result in Talts et al (2018).

@@

This completes one iteration of SBC. To complete the algorithm, multiple iterations are run and the rank statistics are calculated for each parameter. The resulting rank statistics are compared to the discrete uniform distribution to determine if any problematic features exist.

If the computation is well calibrated and the rank statistics follow a discrete uniform distribution, then the posterior credible intervals have sufficient coverage. That is, one way to describe calibration is: for any percentage interval selected over the posterior samples (for example 90%) then there is a 90\% chance that $\theta^{sim}$ falls within this interval. Another way of saying this is a Bayesian analysis is well calibrated if a 90\% credible interval contains the true parameter in 90\% of the SBC iterations.

## 3.2 Implementation
SBC for Hamiltonian Monte Carlo and the Gaussian approximation are performed with 1000 simulation iterations, 1000 observations in each dataset, and post warmup/burnin samples of 999 (which gives 1000 possible rank statistics). This algorithm is summarised below:

@@

## 3.3 Metrics
The key metrics and diagnostics to compare the performance of these methods are the effective sample size (ESS), rank statistics, and chi-squared test statistics.

ESS measures the efficiency of the MCMC sampler. It calculates the number of (effectively) independent draws from the posterior draws generated by a Markov chain. A poor ESS can arise from high autocorrelation in the Markov chain which leads to highly dependent samples. An efficient MCMC algorithm takes less resources (for example, time and number of draws) to get a representative sample of the target distribution. If a MCMC algorithm possesses higher ESS for the majority of its parameters (relative to another strategy), then we may conclude that this method is a more efficient sampler (conditional on the model).

Rank statistics as described in the SBC section are used to evaluate the calibration of a posterior. Histograms will be used to evaluate the distribution of rank statistics. If a posterior is well calibrated then it is expected that the histogram is uniform.

A drawback of this approach is there are more parameters than data points in this model. An alternative to visually checking for uniformity of all the parameters is to calculate the chi squared statistics for the counts in each histogram bin. Let $b_j$ be the number of counts and $e_j$ the expected count in bin $j$. Then the chi squared statistic is given by:

@@

Chi squared statistics can be used to test for uniformity and thus calibration of all the parameters in the model.

@@

# 4. Preliminary Results
Figure 1 shows the distribution of rank statistics for Hamiltonian Monte Carlo, with the horizontal black line at 50 representing the discrete uniform distribution. The preliminary figures focus on the static parameters and arbitrarily chosen states. Phi and sigma_sqd look relatively uniform. However mu and h.995 (the 995th state parameter) have some lumpiness which may suggest a lack of calibration.

@@

A drawback of SBC is that any deviation in uniformity from a calibrated analysis may be due to noisy estimates from small samples. Figure 2 shows more consistent uniform behaviour from all parameters after increasing the number of iterations from 1000 to 5000. This suggests that Hamiltonian Monte Carlo and the overall analysis is well calibrated.

@@

Figure 3 shows results for the Gaussian approximation. The distribution of the rank statistics deviates heavily from the expected discrete uniform distribution, with various peaks at each end of the histogram as well as peaks in the middle. A reason for this miscalibration may be due to the Gaussian mixture only approximating the log chi squared distribution. Hamiltonian Monte Carlo on the other hand, samples directly from the target model which more closely represents the data generating process.

This result may also be due to the relatively small number of posterior samples drawn from the model. The MCMC sampler may not have converged yet which may explain the deviations from uniformity. A follow up step is to compare the ESS of the two methods and to increase the length of the MCMC chain to ensure the ESS is approximately the same before comparing histograms.

![Distribution of parameter ranks for Gaussian mixture model (1000 iterations)](../../simulation_output/sbc_cp_ksc_model_cp_dgf_r1/static_state_hist.png){width=75%}

@@

Kim, Shephard and Chib apply importance sampling weights to correct for the approximation error of the Gaussian mixture model. According to the authors, the correction ensures that samples are from the exact posterior density. The next step is to perform this correction and evaluate the SBC results. I'd expect the Gaussian mixture approximation diagnostics to improve after this correction.

@@

The parameterisation of a model can affect the performance of a sampler. This can cause significant problems in getting unbiased MCMC samples for complex models. An example of this is Neal's funnel (Neal 2003) which arises in hierarchical models which causes performance issues in the Hamiltonian Monte Carlo sampler. Rewriting the model with non centered parameterisation significantly improves the performance of MCMC. Another consideration for this research is to consider both centered and non centered parameterisations of the stochastic volatility model to see if there are any improvements to the simulation diagnostics.

@@

![Model matrix](model_table.jpg)

@@

Week 8: Finish off any simulation or modelling not yet complete (listed in above matrix)

Week 9: Write up results and have a first draft ready for feedback

@@

**Sampling mixture density:**

Rewrite mixture density with respect to a indicator variable $s_t$

$$
\begin{aligned}
&z_t | s_t = i \sim N(m_i - 1.2704, \nu^2) \\
&Pr(s_t = i) = q_i
\end{aligned}
$$

Sample $s_t$ from probability mass function:

$$
\begin{aligned}
Pr(s_t = i | y_t^{\ast}, h_t) \propto q_i f_N(y_t^{\ast} | h_t + m_t - 1.2704, \nu^2)
\end{aligned}
$$

**Sampling steps**

1. Set initial values $s$, $\phi$, $\sigma^2_{\eta}$ and $\mu$

2. Sample $h$ from $h|y^{\ast}, s, \phi, \sigma^2_{\eta}, \mu$

3. Sample s from $s|y^{\ast}, h$

4. Sample $\mu|y^{\ast}, s, \phi, \sigma^2_{\eta}, h$ according to conjugate posterior distribution

4. Sample $\sigma_{\eta}^2|y^{\ast}, s, \phi, \mu, h$ according to conjugate posterior distribution

4. Sample $\phi|y^{\ast}, s, \mu, \sigma^2_{\eta}, h$ using metropolis hastings
